Supplementary Information

### 2021-01-31 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
misc

**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989].** 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
*other global change drivers*
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018].
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].  
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
*(Discuss how this fits with regional to global-scale studies)*
[@martinbenito_convergence_2015]

>>>>

**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989].** 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @babst_twentieth_2019; @walker_integrating_2020].
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018].
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

@@

**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].  
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
*(Discuss how this fits with regional to global-scale studies)*
[@martinbenito_convergence_2015] 
[@babst_twentieth_2019]
